---
title: "lme analysis"
output: html_notebook
---

This document summarizes post-analysis of figuring out what happened in GLMM analysis for the otter drift paper.

Results were saved in an RDS file on 2019-03-13 after running otterDrift_word_Jan2015.Rmd.  This analysis was redone because of reviewers quesioning the use of "month" as a numeric variable, which is a legitimate question. So, I changed the variable to a categorical variable.  

```{r}
rm(list = ls())
library(tidyverse)
library(nlme)

lme.out <- readRDS(file = "RData/results_lme_2019-03-13.rds")
lme.dist <- lme.out$dist
lme.ATOS <- lme.out$ATOS
dat <- lme.out$data
```

First, look at the coefficients of the best model.

```{r}
AICs <- unlist(lapply(lme.dist, 
              FUN = function(x) AIC(x)))

dAIC <- AICs - min(AICs)
lme.dist[AICs == min(AICs)]
```

Look at the relationship between days and distance by ID
```{r}
ggplot(data = dat) +
  geom_point(aes(x = days, y = dist, color = ID)) +
  theme(legend.position = "none")

```

Thinking more... having intercepts is probably unnecessary because every target starts at 0 distance and increases, or not change.  But the results don't change much so ignore this for now... 

```{r}
model_lme <- vector(mode = "list", length = 7)
model_lme[[1]] <- as.formula('dist ~ type * days + wind.spd.sum + cos + sin + month -1')
model_lme[[2]] <- as.formula('dist ~ type * days + cos + sin + month -1')
model_lme[[3]] <- as.formula('dist ~ type + days + wind.spd.sum + cos + sin + month -1')
model_lme[[4]] <- as.formula('dist ~ type + days + cos + sin + month -1')
model_lme[[5]] <- as.formula('dist ~ days + cos + sin + month -1')
model_lme[[6]] <- as.formula('dist ~ wind.spd.sum + cos + sin + month -1')
model_lme[[7]] <- as.formula('dist ~ days + wind.spd.sum + cos + sin + month -1')
model_lme[[8]] <- as.formula('dist ~ type + days -1')
model_lme[[9]] <- as.formula('dist ~ type + days + wind.spd.sum -1')
model_lme[[10]] <- as.formula('dist ~ wind.spd.sum + days -1')
model_lme[[11]] <- as.formula('dist ~ wind.spd.sum + days + month -1')

fit_lme <- lapply(model_lme, 
                  function(x) lme(x,
                                  random = ~ -1 + days|ID, 
                                  data = dat,
                                  control = lmeControl(maxIter = 1000, 
                                                       msMaxIter = 5000,
                                                       niterEM = 5000),
                                  method = "ML"))

varnames <- unlist(lapply(model_lme, 
                          FUN = function(x) strsplit(as.character(x), split = '~')[3]))

AIC_df <- data.frame(AIC = unlist(lapply(fit_lme, 
                                         FUN = function(x) AIC(x))),
                     model = varnames) 
AIC_df$delta <- AIC_df$AIC - min(AIC_df$AIC)
rownames(AIC_df) <- paste("Model", 1:length(model_lme)) 
#model.lkhd <- exp(-0.5 * AIC_df$delta)
#AIC_df$weight <- model.lkhd/(sum(model.lkhd))

AIC.ordered <- AIC_df[order(AIC_df$delta), 
                      colnames(AIC_df) != "AIC"]

```


Back to the original issue of the effects of month on distance traveled, where all coefficients are negative except for February
```{r}
coef(fit_lme[[11]])
```

Let's take a look at how many individual targets were found within each month.

```{r}
dat %>% group_by(month) %>% summarise(count = n_distinct(ID))

```

Unequal sample sizes for sure... the negative coefficients come from the decreasing distances within a month.  Is there a correlation between month and cumulative wind speed?  So, take a look at how distances changes within a month per ID - just the means:

```{r}
dat %>% group_by(month, ID) %>% summarise(mean.dist = mean(dist), n = n()) -> tmp

ggplot(data = tmp) +
  geom_point(aes(x = month, y = mean.dist, color = ID),
             size = 2) +
  theme(legend.position = "none")
```

The raw data show that winter has large values whereas summer has less - except January. 

```{r}
ggplot(data = dat) +
  geom_point(aes(x = month, y = dist, color = ID),
             size = 2) + 
  theme(legend.position = "none")
```

Take a look at monthly sample sizes
```{r}
dat %>% group_by(month) %>% summarize(n.data = n(), 
                                      n.ID = n_distinct(ID),
                                      mean.dist = mean(dist))
```

I don't know why those coefficients are negative... 

```{r}

M11 <- as.formula('dist ~ wind.spd.sum + days + month')
fit <- lme(M11,
           random = ~ 1 + days|ID, 
           data = dat,
           control = lmeControl(maxIter = 1000, 
                                msMaxIter = 5000,
                                niterEM = 5000),
           method = "ML")

summary(fit)
```

Take a look at coefficients:

```{r}
coefs <- coef(fit)
```

These coefficients can be used as follows to create predicted values:

```{r}
IDs <- rownames(coefs)
k <- 1
out.list <- vector(mode = "list", length = length(IDs))
for (k in 1:length(IDs)){
  # get the data for each indiv
  tmp <- filter(dat, ID == IDs[k])
  # create design matrix. Need to look at tmp to figure out "month" part of zeros and ones. Note 
  # there is no month1, which is in intercept.
  X.month <- matrix(data = 0, nrow = nrow(tmp), ncol = 11)
  tmp.month <- cbind(seq(1, nrow(tmp)), tmp$month)
  X.month[tmp.month] <- 1
  
  design.X <- cbind(1, tmp$wind.spd.sum, 
                    tmp$days, X.month[, 2:11])
  
  # when multiply this with the coefficients, I get the regression line for the individual:
  y.hat <- design.X %*% t(coefs[k,])
  out.list[[k]] <- data.frame(ID = IDs[k],
                              days = tmp$days,
                              y = tmp$dist,
                              yhat = as.vector(y.hat))
}

out.df <- do.call(rbind, out.list)

ggplot(data = out.df) + 
  geom_point(aes(x = days, y = y, color = ID)) + 
  geom_path(aes(x = days, y = yhat, color = ID)) + 
  theme(legend.position = "none")
# Here are the observed values
#tmp$dist

# these should be able to be plotted - a line per individual... 
# Let's do that soon. 

```

