---
title: "Otter Drift Experiment"
author: "Tomo Eguchi"
date: '`r Sys.Date()`'
output: word_document
fontsize: 12pt
linestretch: 2
header-includes:
  - \usepackage{lineno}
  - \linenumbers
  - \setlength\parindent{24pt}
---

```{r global_options, include=FALSE}
#'`r Sys.Date()`'
#date: \today{}   # works only for pdf output

knitr::opts_chunk$set(fig.path='figures/', echo = FALSE, message = FALSE)
```

```{r getStarted, include = FALSE}
library(nlme)
library(gamm4)
library(knitr)
library(ggplot2)
rm(list=ls())
runDate <- Sys.Date()
sysInfo <- Sys.info()
ifelse(sysInfo[1] == 'Linux',
       source('~/Documents/R/TomosFunctions.R'),
       source('~/R/TomosFunctions.R'))

datafile <- 'Data_14Feb2014a.csv'

# the following two are used to count figure and table #s 
figCount <- c(`_` = 0)
tableCount <- c(`_` = 0)

```

```{r manipulateData}
dat <- read.table(file = datafile, sep=",", header = TRUE)
#windDat <- read.table(file = windfile, sep = ",", header = TRUE)

dat_M <- vector(mode = "numeric", length = dim(dat)[1])
dat_Y <- vector(mode = "numeric", length = dim(dat)[1])
for (k in 1:dim(dat)[1]){
  dat_M[k] <- as.numeric(strsplit(as.character(dat$Date[k]), "/")[[1]][1])
  dat_Y[k] <- as.numeric(strsplit(as.character(dat$Date[k]), "/")[[1]][3])
}

dat$dat_M <- dat_M
dat$dat_Y <- dat_Y - min(dat_Y)

dat3_mat <- dat[,c("DaysSinceRelease", "windSpd", "windDir", 
                   "windSpdSum", "windDirSum", "dat_M")]
dat3_mat_std <- scale(dat3_mat)
dat3_df <- as.data.frame(cbind(dat3_mat_std, 
                               dat[, c("Type", "distInKm", "dat_Y", "ID")]))
```

#Methods
We considered physical variables that possibly affected the drift distance of carcasses and dummies. These variables were wind speed, wind directions, cumulative wind speed, and cumulative wind directions. For each carcass or dummy, a total distance traveled was computed at every location fix as well as the physical variables. We first fitted linear models to the data to determine (1) whether drift distances were different between carcasses and dummies and (2) what variables affect the drift distance. Because of the repeated measurements of each target (carcass/dummy) over time, we treated the each taget as a random variable and used the mixed effects models. All predictor variables were centered and standardized. After finding useful predictors, we fitted a generalized additive model with cubic regression spline as the basis function to the data to determine if a non-linear model might be more appropriate. The upper bound of the basis function was set at 5, which limited excessive wiggliness. All analyses were conducted using the R statistical environment (V. 3.2.1; R Core Team 2015) with packages nlme (Pinheiro et al. 2015) and gamm4 (Wood and Scheipl 2014). 

```{r linearModels}
model_lme <- vector(mode = "list", length = 3)
model_lme[[1]] <- as.formula('distInKm ~ Type + DaysSinceRelease + windSpdSum + windDirSum')
model_lme[[2]] <- as.formula('distInKm ~ Type + DaysSinceRelease + windDirSum')
model_lme[[3]] <- as.formula('distInKm ~ DaysSinceRelease + windDirSum')

fit_lme <- lapply(model_lme, 
                  function(x) lme(x,
                                  random = ~ 1 + DaysSinceRelease|ID, 
                                  data = dat3_df,
                                  control = lmeControl(maxIter = 1000, 
                                                       msMaxIter = 5000,
                                                       niterEM = 5000),
                                  method = "ML"))

#summary(fit_lme3)
varnames <- unlist(lapply(model_lme, 
                          FUN = function(x) strsplit(as.character(x), split = '~')[3]))

AIC_df <- data.frame(AIC = unlist(lapply(fit_lme, 
                                         FUN = function(x) AIC(x))),
                     model = varnames) 
AIC_df$delta <- AIC_df$AIC - min(AIC_df$AIC)
rownames(AIC_df) <- c("Model 1", "Model 2", "Model 3")
model.lkhd <- exp(-0.5 * AIC_df$delta)
AIC_df$weight <- model.lkhd/(sum(model.lkhd))

AIC.ordered <- AIC_df[order(AIC_df$delta), colnames(AIC_df) != "AIC"]
tableCount <- incCount(tableCount, "table_AIC")
text_table_AIC <- "A comparison of linear models fitted to drift data of sea otter carcasses and dummies. delta is difference in AIC values and Weight is Akaike weight (Burnham and Anderson 2010)."

bestModel <- rownames(AIC.ordered)[1]
bestModelNum <- as.numeric(unlist(strsplit(bestModel, split = "Model "))[2])
#AIC(fit_lme1, fit_lme2, fit_lme3)

coefTable <- as.data.frame(summary(fit_lme[[bestModelNum]])$tTable)
coefTable <- coefTable[, (colnames(coefTable) != "DF" &  
                            colnames(coefTable) != "t-value")]

tableCount <- incCount(tableCount, "table_coefBestModel")
text_table_coefBestModel <- "Estiamted coefficients of the best model."
```

```{r gamModel, include=FALSE}
fit2_gam <- gamm4(distInKm ~ s(DaysSinceRelease, bs = "cr", k = 5, by = Type) + 
                    s(windDirSum, bs = "cr",k = 5, by = Type),
                  random = ~(1 + DaysSinceRelease|ID),
                  data = dat3_df,
                  family = gaussian)

plot_gam <- plot(fit2_gam$gam)

# make plots for all combos
plotAll_GAM <- lapply(plot_gam, 
                      FUN = function(y){
                        df_gam <- data.frame(v1 = y$x,
                                             smooth = y$fit,
                                             SE = y$se)
                        df_rug <- data.frame(v1 = y$raw)
                        
                        plot1_GAM <- ggplot() + 
                          geom_ribbon(aes(x=v1, 
                                          y=smooth, 
                                          ymin = smooth-1.96*SE, 
                                          ymax=smooth+1.96*SE), 
                                      data = df_gam) + 
                          geom_line(aes(x=v1,
                                        y=smooth), 
                                    data = df_gam, 
                                    color = "white") + 
                          xlab(y$xlab) + 
                          ylab(y$ylab) + 
                          theme(axis.text = element_text(size = 12)) + 
                          geom_rug(data=df_rug, aes(x=v1)) + 
                          theme_minimal()} )


figCount <- incCount(figCount, "figure_plotGam")

text_plotGam <- "The GAM smooth functions for days since release and cumulative sum of wind directions as predictors to distance traveled of sea otters.  Black ribbons indicate approximate 95% confidence intervals. Variables were centered and standardized prior to the anlaysis."

```

#Results
Among the three linear models, `r bestModel` had the smallest AIC value, although the delta AIC values were less than 4 for the other two (Table `r labelNumber(tableCount, "table_AIC")`). However, because the estimated coefficients for the common predictors (i.e., DaysSinceRelease and windDirSum) from the other two models were virtually identical to those from the best model (Table `r labelNumber(tableCount, "table_coefBestModel")`; DaysSinceRelease: `r signif(summary(fit_lme[[1]])$tTable["DaysSinceRelease", "Value"], digits = 4)` and `r signif(summary(fit_lme[[2]])$tTable["DaysSinceRelease", "Value"], digits = 4)`, windDirSum: `r signif(summary(fit_lme[[1]])$tTable["windDirSum", "Value"], digits = 4)` and `r signif(summary(fit_lme[[2]])$tTable["windDirSum", "Value"], digits = 4)`), we did not use the model averaging. 

\noindent `r pasteLabel(preText = "Table", inObj = tableCount, objName = "table_AIC", postText = text_table_AIC)`
`r kable(AIC.ordered, digits = 2)`

Drifting distances were not statistically different between the carcasses and dummies.  
In other words, carcasses and dummies drifted similar distances. Drifting distances were positively affected with time but negatively affected by the cumulative direction of wind. The cumulative direction indicates ... 

\noindent `r pasteLabel(preText = "Table", inObj = tableCount, objName = "table_coefBestModel", postText = text_table_coefBestModel)`
`r kable(coefTable, digits = 2)`

To consider non-linear relationships between the predictors and distance moved, a generalized additive model also was fitted to the data. For this approach, `r bestModel` was fitted to carcasses and dummies, where cubic regression spline was used as the basis function. Relationshis could be approximated by linear functions because of wide uncertainty and small estimated degrees of freedom (< 3; Figure `r labelNumber(figCount, "figure_plotGam")`). 


```{r plotGamFunctions}
multiplot(plotlist = plotAll_GAM, cols = 2)
```

\noindent `r pasteLabel(preText = "Figure", inObj = figCount, objName = "figure_plotGam", postText = text_plotGam)`



